Learning Discrete Bayesian Networks from Continuous Data

### 2015-12-01 BN_discretizer_paper.tex
Finish the update

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
%  gsave
%  newpath
%    20 20 moveto
%    20 220 lineto
%    220 220 lineto
%    220 20 lineto
%  closepath
%  2 setlinewidth
%  gsave
%    .4 setgray fill
%  grestore
%  stroke
%  grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usepgfplotslibrary{groupplots}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{boondox-cal} % for lowercase caligraphic
\usepackage[font=small,skip=5pt]{caption}

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning and inference algorithms based on the parameterizations.
One successful example of parametric, continuous distributions in Bayesian networks is the Gaussian graphical model \citep{Weiss_2011}. \yichun{Is the comma between parametric and continuous necessay?}
%For other parametric distributions and non-linear functions \citep{Ihler_2009}, however, these redesigned algorithms can be computationally expensive.

@@

\yichun{A common discretization method for Bayesian networks is to split continuous variables with uniform width or to decide discretization edges based on expert knowledge. A more principled method is the minimum description length (MDL) principle discretization \citep{Friedman_1996}.}
The MDL principle proposed by \cite{MDL_1978} states that the best model for a dataset is the one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description length of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.

@@

Let $D_{\bm{Y}}$ be the subset of the training data corresponding to variable subset $\bm{Y}$.
Four principles for the optimal discretization policy enable the formulation of $P(\Lambda)$ and $P(D_{-\cX} \mid \Lambda)$, where $D_{-\cX}$ is the subset of the dataset for all variables but $\cX$ and the probability of the data associated with the target variable, $P(D_{\cX})$, is already captured in $P(\Lambda)$.
The four principles,\yichun{ which can be considered an extension of the priors in MODL and \cite{Lustgarten_2011} to Bayesian networks}, are:

@@

The procedure used to minimize the objective function involves \yichun{dynamical programming}.
Note that because the objective function is cumulative over intervals, if a partition of $\cX$ into $k$ intervals by $\Lambda = \anglebrackets{\gamma_1, \gamma_2, \ldots, \gamma_k}$ is an optimal discretization policy, then any subinterval is optimal for the corresponding subproblem.
It follows that dynamic programming can be used to solve the optimization problem exactly.

@@

Two experiments were conducted on each dataset.
The first experiment compares the performance of the Bayesian and MDL discretization methods on a known Bayesian network structure.
The structure was obtained by prediscretizing each continuous variable into $k$ uniform-width intervals, where $k$ is the median number of instantiations of the discrete variables, and using the structure with the highest Bayesian score from \num{1000} runs of the K2 algorithm with random topological orderings.
The second experiment compares the same methods applied when simultaneously discretizing and learning network structure.

@@

\begin{equation}
  \label{eqn:MDL}
  \begin{aligned}
  \frac{1}{2} \ln(n) \left\lbrace  \lvert \Pa_{\cX} \rvert (\lvert X \rvert - 1) +
   {\sum_{j,\cX \in \Pa_{X_j}}} \lvert \Pa_{X_j} \rvert (\lvert X_j \rvert - 1) \right\rbrace + \ln(|X|) \\
   + (m-1) H \paren{\frac{|X| - 1}{m -1}} -n \cdot \left[ I(X,\Pa_{\cX}) + {\sum_{j,\cX \in \Pa_{X_j}}} I(X_j, \Pa_{X_j}) \right]\text{,}
  \end{aligned}
\end{equation}
where $I(A,B)$ is the mutual information between two discrete variable sets $A$ and $B$, ${H(p) = -p \ln(p) - (1-p) \ln(1-p)}$, and $X$ is the discretized version of $\cX$.
Equation \ref{eqn:MDL} can be minimized using dynamic programming with runtime $\bigo \paren{ n^3 + \paren{ n_c \cdot {L}^{n_s}  + {L}^{n_p}} \cdot n^2 }$.
All variables are defined in Section \ref{subsec:algo}.
Therefore, the runtime of the proposed method in this paper is shorter than the MDL discretization method by $\bigo \paren{n^3}$.

@@

\begin{table}[ht]
  \centering
  \caption{
    Results from discretization of the Auto MPG dataset with fixed structure from Figure~\ref{fig:auto_graph_1}.
    The first five rows list the discretization edges for each continuous variable.
    The last row lists the mean cross-validated log-likelihood; positive values are better.
    The MDL method does not produce any discretization edges.
  }
  \input{table_autoMPG_1.tex}
  \label{table:auto_disc_table_1}
\end{table}

@@

\begin{table}
  \centering
  \caption{
    Discretization policy summary of the Housing dataset based on the fixed network structure shown.
    The first twelve rows show the numbers of intervals after discretization for each method.
    The last row is the mean cross-validated log-likelihood based on each discretization result.
    % The heuristic method is over-sensitive and discretizes some variables into too many intervals.
    % The MDL method is less sensitive and on many cases does not produce any discretization intervals.
    % The optimal Bayesian approach produces a reasonable number of discretization edges.
  }
  \input{table_housing_1.tex}
  \label{table:housing_disc_table_1}
\end{table}

@@

\begin{figure}[H]
  \centering
  \input{graph_housing_exp1_distr_8_5}
  \caption{
    The discretization policies for variables $H$ and $E$ learned using the optimal Bayesian approach on the fixed network shown in Figure~\ref{fig:housing_graph_1}.
    The variable $H$ possesses no \todo{overly} repeated values, thus producing a cleaner discretization policy.
  }
  \label{fig:housing_exp1_distr_8_5}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
This paper introduced a principled discretization method for continuous variables in Bayesian networks with quadratic complexity instead of the cubic complexity of other standard techniques.
Emperical demonstrations show that the proposed method is superior to the state of the art.
In addition, this paper shows how to incorporate existing methods into the structure learning process to discretize all continuous variables and simultaneously learn Bayesian network structures.
The proposed method was incorporated and its superior performance was empirically demonstrated.
All software is publically available to support discretization research, \todo{and can be found at.}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
%\begin{filecontents*}{example.eps}
  %!PS-Adobe-3.0 EPSF-3.0
  %%BoundingBox: 19 19 221 221
  %%CreationDate: Mon Sep 29 1997
  %%Creator: programmed by hand (JK)
  %%EndComments
%  gsave
%  newpath
%    20 20 moveto
%    20 220 lineto
%    220 220 lineto
%    220 20 lineto
%  closepath
%  2 setlinewidth
%  gsave
%    .4 setgray fill
%  grestore
%  stroke
%  grestore
%\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{pgfplotstable}
\usepackage{tikz}
\usepgfplotslibrary{groupplots}
\usepackage{lscape}
\usepackage{caption}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{boondox-cal} % for lowercase caligraphic
\usepackage[font=small,skip=5pt]{caption}
\usepackage[hyphens]{url}

@@

There are three common approaches to extending Bayesian networks to continuous variables.
The first is to model the conditional probability density of each continuous variable using specific families of parametric distributions, and then to redesign Bayesian network learning and inference algorithms based on the parameterizations.
One successful example of parametric, continuous distributions in Bayesian networks is the Gaussian graphical model \citep{Weiss_2011}. \yichun{Is the comma between ``parametric" and ``continuous" necessay?}
%For other parametric distributions and non-linear functions \citep{Ihler_2009}, however, these redesigned algorithms can be computationally expensive.

@@

\yichun{A common discretization method for Bayesian networks is to split continuous variables by uniform width or by expertises from application fields. A more principled method is the minimum description length (MDL) principle discretization \citep{Friedman_1996}.}
The MDL principle proposed by \cite{MDL_1978} states that the best model for a dataset is the one that minimizes the amount of information needed to describe it \citep{Grunwald_2009}.
MDL methods trade off goodness-of-fit against model complexity to reduce generalization error.
In the context of Bayesian networks, \cite{Friedman_1996} applied the MDL principle to determine the optimal number of discretization intervals for continuous variables and the optimal positions of their discretization edges.
Their approach selects a discretization policy that minimizes the sum of the description length of the discretized Bayesian network and the information necessary for recovering the continuous values from the discretized data.

@@

Let $D_{\bm{Y}}$ be the subset of the training data corresponding to variable subset $\bm{Y}$.
Four principles for the optimal discretization policy enable the formulation of $P(\Lambda)$ and $P(D_{-\cX} \mid \Lambda)$, where $D_{-\cX}$ is the subset of the dataset for all variables but $\cX$ and the probability of the data associated with the target variable, $P(D_{\cX})$, is already captured in $P(\Lambda)$.
The four principles, \yichun{ which can be considered an extension of the priors in MODL and \cite{Lustgarten_2011} to Bayesian networks}, are:

@@

The procedure used to minimize the objective function involves dynamical programming.
Note that because the objective function is cumulative over intervals, if a partition $\Lambda = \anglebrackets{\gamma_1, \gamma_2, \ldots, \gamma_k}$ of $\cX$ is an optimal discretization policy, then any subinterval is optimal for the corresponding subproblem.
It follows that dynamic programming can be used to solve the optimization problem exactly.

@@

Two experiments were conducted on each dataset.
The first experiment compares the performance of the Bayesian and MDL discretization methods on a known Bayesian network structure.
The structure was obtained by discretizing each continuous variable into $k$ uniform-width intervals, where $k$ is the median number of instantiations of the discrete variables, and using the structure with the highest Bayesian score from \num{1000} runs of the K2 algorithm with random topological orderings.
The second experiment compares the same methods applied when simultaneously discretizing and learning network structure.

@@

\begin{equation}
  \label{eqn:MDL}
  \begin{aligned}
  \frac{1}{2} \ln(n) \left\lbrace  \lvert \Pa_{\cX} \rvert (\lvert X \rvert - 1) +
   {\sum_{j,\cX \in \Pa_{X_j}}} \lvert \Pa_{X_j} \rvert (\lvert X_j \rvert - 1) \right\rbrace + \ln(|X|) \\
   + (m-1) H \paren{\frac{|X| - 1}{m -1}} -n \cdot \left[ I(X,\Pa_{\cX}) + {\sum_{j,\cX \in \Pa_{X_j}}} I(X_j, \Pa_{X_j}) \right]\text{,}
  \end{aligned}
\end{equation}
where $I(A,B)$ is the mutual information between two discrete variable sets $A$ and $B$, ${H(p) = -p \ln(p) - (1-p) \ln(1-p)}$, and $X$ is the discretized version of $\cX$.

\yichun{Equation \ref{eqn:MDL} can be minimized exactly using dynamic programming. For a given Bayesian network, the first three terms only depends on the number of discretization intervals $\lvert X \rvert$ and the fourth term is cumulative on the intervals. Hence, if $\Lambda = \anglebrackets{\gamma_1, \gamma_2, \ldots, \gamma_k}$ is a MDL optimal discretization policy with $k$ intervals, then  $\Lambda' = \anglebrackets{\gamma_1, \gamma_2, \ldots, \gamma_{k-1}}$ is a MDL optimal discretization policy of the corresponding subproblems with $k-1$ intervals.}
This procedure takes runtime $\bigo \paren{ n^3 + \paren{ n_c \cdot {L}^{n_s}  + {L}^{n_p}} \cdot n^2 }$. 
All variables are defined in Section \ref{subsec:algo}.
Therefore, the runtime of the proposed method in this paper is shorter than the MDL discretization method by $\bigo \paren{n^3}$, since the former has a more efficient form of dynamical programming.

@@

\begin{table}[ht]
  \centering
  \caption{
    Discretization result of the Auto MPG dataset with fixed structure from Figure~\ref{fig:auto_graph_1}.
    The first five rows list the discretization edges and the last row lists the mean cross-validated log-likelihood; positive values are better.
    %The MDL method does not produce any discretization edges.
  }
  \input{table_autoMPG_1.tex}
  \label{table:auto_disc_table_1}
\end{table}

@@

\begin{table}
  \centering
  \caption{
    Discretization policy summary of the Housing dataset based on the fixed network structure shown.
    The first twelve rows show the numbers of discretization intervals and the last row is the mean cross-validated log-likelihood.% based on each discretization result.
    % The heuristic method is over-sensitive and discretizes some variables into too many intervals.
    % The MDL method is less sensitive and on many cases does not produce any discretization intervals.
    % The optimal Bayesian approach produces a reasonable number of discretization edges.
  }
  \input{table_housing_1.tex}
  \label{table:housing_disc_table_1}
\end{table}

@@

\begin{figure}[H]
  \centering
  \input{graph_housing_exp1_distr_8_5}
  \caption{
    The discretization policies for variables $H$ and $E$ learned using the optimal Bayesian approach on the fixed network shown in Figure~\ref{fig:housing_graph_1}.
    \yichun{The variable $H$ possesses less frequenly repeated values, thus producing less number of discretization intervals than $E$.}
  }
  \label{fig:housing_exp1_distr_8_5}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
This paper introduced a principled discretization method for continuous variables in Bayesian networks with quadratic complexity instead of the cubic complexity of other standard techniques.
Empirical demonstrations show that the proposed method is superior to the state of the art.
In addition, this paper shows how to incorporate existing methods into the structure learning process to discretize all continuous variables and simultaneously learn Bayesian network structures.
The proposed method was incorporated and its superior performance was empirically demonstrated.
All software is publically available to support discretization research, and \yichun{can be found at \url{github.com/sisl/LearnDiscreteBayesNets.jl}}.
